Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-06 paper2.qmd
section:model; process:write,figures

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons where a single judge evaluates multiple stimuli [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the DP for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp. 266]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

@@

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts. The figure indicates that the mDP for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality. Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces *the law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as *the discriminal difference*, determines the observed dichotomous outcome. Moreover, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted that the general form of the theory, as delineated in @sec-thurstone_theory, gave rise to a trait scaling problem. The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each case progressively incorporated additional simplifying assumptions into the model.

@@

<!-- 3. The Case V and its simpler statistical representation: the BTL model -->
Despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16].

@@

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, this process occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp. 150], particularly the assumption of equal dispersions. For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of these discriminal difference ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.

@@

<!-- 4. So, what happens when the assumption does not hold? -->
Nevertheless, model misspecification, in the form of an erroneous assumption of equal dispersions between stimuli, can give rise to significant statistical and measurement issues. For instance, the model may overestimate the degree to which the outcome accurately reflects the "true" discriminal differences between stimuli. This overestimation can result in researchers drawing spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.

Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information. Such exclusions can introduce bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
The correlation, represented by the symbol $\rho$, measures how much a judge's perception of a specific trait in one stimulus depends on their perception of the same trait in another. As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons. @fig-correlation presents a similar thought experiment as in @sec-theory-issue1a to illustrate this idea. The illustration now assumes that the discriminal dispersions for both texts remain constant. The figure shows that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower. This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor Text B over Text A (shaded gray area) compared to scenarios where the texts are highly correlated.

@@

<!-- 2. But this is not the scenario we observe -->
Off course, in experimental practice, researchers approach this process in reverse. They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli. Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, then the model estimates a discriminal difference distribution that accurately reflects the "true" discriminal difference of the texts. This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. Once more, the estimation accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.

@@

<!-- 5. But what are the challenges when this assumption is violated? -->
In any case, similar to @sec-theory-issue1a, model misspecification due to an erroneous assumption of zero correlation between stimuli can lead to significant statistical and measurement issues. For instance, the model may over- or underestimate how accurately the outcome reflects the "true" discriminal differences between stimuli. Such inaccuracies can result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes. This scenario is illustrated in @fig-correlation, where the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.

This misspecification may arise from neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures. Neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp. 341] or, worse, introducing bias into the trait's estimates [@Ackerman_1989]. Excluding judges based on misfit statistics risks discarding valuable information, which may further bias the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Finally, ignoring hierarchical structures may reduce the precision of model parameter estimates, potentially amplifying the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp. 482].

@@

This section presents a theoretical model for CJ that extends Thurstone's theory. It uses causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014; @Hernan_et_al_2020] and, in particular, directed acyclic graphs (DAGs) [@Gross_et_al_2018; @Neal_2020], to articulate the core theoretical principles of CJ theory. The model also incorporates several practical factors that influence judges in CJ experiments, such as the selection of judges, stimuli, and the number of comparisons. In addition, the study uses Bayesian statistical methods [@McElreath_2020] to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.

@@

::: {.panel-tabset group="Figure"}

#### Figure

#### 3

![](/images/png/CJ_TM_A1.png){#fig-CJ_TM_A1 width=20%}

#### 4

![](/images/png/CJ_TM_A2.png){#fig-CJ_TM_A2 width=30%}

#### 5

![](/images/png/CJ_TM_A3.png){#fig-CJ_TM_A3 width=42%}

#### 6

![](/images/png/CJ_TM_A4.png){#fig-CJ_TM_A4 width=42%}

@@

::: panel-tabset

#### Figure

#### 7

![](/images/png/CJ_TM_B1.png){#fig-CJ_TM_B1 width=42%}

#### 8

![](/images/png/CJ_TM_B2.png){#fig-CJ_TM_B2 width=62%}

#### 9

![](/images/png/CJ_TM_B3.png){#fig-CJ_TM_B3 width=72%}

#### 10

![](/images/png/CJ_TM_B4.png){#fig-CJ_TM_B4 width=72%}

#### 11

![](/images/png/CJ_TM_B5.png){#fig-CJ_TM_B5 width=72%}

#### 12

![](/images/png/CJ_TM_B6.png){#fig-CJ_TM_B6 width=72%}

Considering the sampling and comparison mechanisms

::: panel-tabset

#### Figure

#### 13

![](/images/png/CJ_TM_B7.png){#fig-CJ_TM_B7 width=90%}

#### 14

![](/images/png/CJ_TM_B8.png){#fig-CJ_TM_B8 width=100%}

#### 15

![](/images/png/CJ_TM_B9.png){#fig-CJ_TM_B9 width=100%}

>>>>

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons where a single judge evaluates multiple stimuli [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp. 266]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

@@

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts. The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality. Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Moreover, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion is supported by two key observations: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that the general form of the theory, as outlined in @sec-thurstone_theory, gave rise to a problem of trait scaling. The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each case progressively incorporated additional simplifying assumptions into the model.

@@

<!-- 3. The Case V and its simpler statistical representation: the BTL model -->
Notably, despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16].

@@

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, this process occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Therefore, the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp. 150], particularly the assumption of equal dispersions. For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of these discriminal difference ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.

@@

<!-- 4. So, what happens when the assumption does not hold? -->
Thus, model misspecification, in the form of an erroneous assumption of equal dispersions between stimuli, can give rise to significant statistical and measurement issues. For instance, the model may overestimate the degree to which the outcome accurately reflects the "true" discriminal differences between stimuli. This overestimation can result in researchers drawing spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information. Such exclusions can introduce bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
The correlation, represented by the symbol $\rho$, measures how much a judge's perception of a specific trait in one stimulus depends on their perception of the same trait in another. As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons. @fig-correlation presents a similar thought experiment as in @sec-theory-issue1a to illustrate this idea. The illustration now assumes that the discriminal dispersions for both texts remain constant. The figure reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower. This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor Text B over Text A (shaded gray area) compared to scenarios where the texts are highly correlated.

@@

<!-- 2. But this is not the scenario we observe -->
Off course, in experimental practice, researchers approach this process in reverse. They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli. Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, then the model estimates a discriminal difference distribution that accurately reflects the "true" discriminal difference of the texts. This scenario is also illustrated in @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. Once more, the estimation accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.

@@

<!-- 5. But what are the challenges when this assumption is violated? -->
In any case, similar to @sec-theory-issue1a, model misspecification due to an erroneous assumption of zero correlation between stimuli can lead to significant statistical and measurement issues. For instance, the model may over- or underestimate how accurately the outcome reflects the "true" discriminal differences between stimuli. Such inaccuracies can result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes. This scenario is also illustrated by @fig-correlation, when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.

The misspecification may arise from neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures. Neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp. 341] or, worse, introducing bias into the trait's estimates [@Ackerman_1989]. Excluding judges based on misfit statistics risks discarding valuable information, which may further bias the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Finally, ignoring hierarchical structures may reduce the precision of model parameter estimates, potentially amplifying the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp. 482].

@@

This section presents a theoretical model for CJ that extends Thurstone's theory. It uses causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014] and, in particular, directed acyclic graphs (DAGs) [@Gross_et_al_2018; @Neal_2020], to articulate the core theoretical principles of CJ theory. The model also incorporates several practical factors that influence judges in CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition, the study uses Bayesian statistical methods [@McElreath_2020] to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.

@@

<!-- ![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%} -->

<!-- ![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%} -->

<!-- ![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%} -->

![](/images/png/CJ_TM_04.png){#fig-CJ_TM_04 width=45%}

@@

<!-- ![](/images/png/CJ_TM_05.png){#fig-CJ_TM_05 width=44%} -->

<!-- ![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%} -->

<!-- ![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%} -->

<!-- ![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%} -->

<!-- ![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%} -->

![](/images/png/CJ_TM_10.png){#fig-CJ_TM_10 width=68%}

Considering the sampling mechanism

<!-- ![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%} -->

![](/images/png/CJ_TM_12.png){#fig-CJ_TM_12 width=80%}

Considering comparison mechanisms

<!-- ![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%} -->

![](/images/png/CJ_TM_14.png){#fig-CJ_TM_14 width=90%}

<!-- ![](/images/png/CJ_TM_15.png){#fig-CJ_TM_15 width=90%} -->
